---
title: "Problem Set 3"
author: "Ted Dunmire"
date: "Saturday, June 20, 2015"
output: html_document
---


1. Effects of Facebooks ads

    ```{r}
    fbData <- read.csv("C:/Users/tdunmire/Downloads/broockman_green_anon_pooled_fb_users_only.csv")
    ```

    a. Using regression without clustered standard errors compute confidence interval for the effect of the ad on candidate name recognition (Study 1 only)
    
    ```{r}
    modela <- lm(name_recall ~ treat_ad, data = fbData[fbData$studyno == 1,])
    summary(modela)
    
    #now get the standard error
    stanErr <- summary(modela)$coefficients[2,2]
    marErr <- stanErr * 1.96
    point_estimate <- summary(modela)$coefficients[2,1]
    lower <- point_estimate - marErr
    upper <- point_estimate + marErr
    ```
    
    So this will indicate the 95% confidence interval is between **`r lower`** and **`r upper`**
    
    b. What are the clusters in this study? Why might taking clustering into account increase the standard errors
    
    The clustering in this experiment are specific combinations of age, gender, and location. Pg 82-83 GG                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 
    
    c. Now repeat part (a) but taking clustering into account: compute a confidence interval for the effect of the ad on candidate name recognition in STudy 1 but now account for cluster
    
    ```{r}
    #Cluster function provided in instructions
    cl <- function(fm, cluster){
    require(sandwich, quietly = TRUE)
	  require(lmtest, quietly = TRUE)
	  M <- length(unique(cluster))
	  N <- length(cluster)
	  K <- fm$rank
	  dfc <- (M/(M-1))*((N-1)/(N-K))
	  uj <- apply(estfun(fm),2, function(x) tapply(x, cluster, sum));
	  vcovCL <- dfc*sandwich(fm, meat=crossprod(uj)/N)
	  coeftest(fm, vcovCL)
    }
    
    study1 <- fbData[fbData$studyno == 1,] 
    study1$cluster <- factor(as.character(study1$cluster))
    modelb <- lm(name_recall ~ treat_ad, data = study1)
    resultc <- cl(modelb, study1$cluster) #clean up and make look nice
    ```
    
    ***Returned Standard Error = NA???***
    
    d. Repeat part (c) but now for study 2 only
    
    ```{r}
    study2 <- subset(fbData, fbData$studyno == 2 & fbData$name_recall != 'NA')
    study2$cluster <- factor(as.character(study2$cluster))
    modeld <- lm(name_recall ~ treat_ad, data = study2)
    resultd <- cl(modeld, study2$cluster) #arguments must have same length 
    #[-0.0723899, 0.0667832]
    ```
    
    ***TBD based on answer above***
    
    e. Repeat part (c) but using the entire sample. What is the treatment effect and associated p-value?
    
    ```{r}
    study3 <- subset(fbData, fbData$name_recall != 'NA')
    study3$cluster <- as.factor(as.character(study3$cluster))
    modele <- lm(name_recall ~ treat_ad, data = study3)
    resulte <- cl(modele, study3$cluster)
    #-0.207465, -0.1026815
    ```
    
    ***TBD based on answer above***
    
    f. Repeat part e but include a dummy variable for whether data is from study 1 or study 2
    
    ```{r}
    fbData2 <- fbData
    fbData2$dummy <- fbData2$studyno - 1
    
    study4 <- subset(fbData2, fbData2$name_recall != 'NA')
    study4$cluster <- as.factor(as.character(study4$cluster))
    modelf <- lm(name_recall ~ treat_ad + dummy, data = study4)
    resultf <- cl(modelf, study4$cluster) #different standard errors?
    ```
    
    g. Why are the results from e and f different. Which result is biased and why (see pg 75-76)? 
    
    Simulating a blocked assignment 
    The result in part e is biased because there is not an equal chance of being assigned to treatment across both groups. 
    
    h. Skim Facebook study and consider two claims made, why might those be different
    
    i. There was a 19% difference in the way people voted in areas where Facebook Ads ran vs areas where ads did not run. 
    
    ii. In ares where ads ran, people with the most online exposure were 17 percent more likely to vote against the proposition. 
    
    
    ***Figure this out***
    
    
2. Read article on encouraging recycling in Peru ***Cannot access journal***

3. Questions on Multi-factor design of the recycling experiment
